```{r postblog, eval=FALSE, echo=FALSE}
## Run these commands when you are ready to push out to the fishR blog.
title <- "R-Squared for a VBGM"
cats <- c('Growth','Statistics')
keys <- c('Growth','von Bertalanffy')
setwd("C:/aaaWork/Web/fishR/BlogExamples")
source('WPSetup.R')
knit2wp("nlsRSquared.Rmd",publish=FALSE,shortcode=TRUE,
        title=title,categories=cats,mt_keywords=keys)
#NOTE that I had to manually edit the equations in wordpress.
```

```{r setup, echo=FALSE, results='hide'}
# Having this separate from the setup file allows one to knit this in RStudio.
opts_chunk$set(dev="png",tidy=FALSE)
```

Recently, a fishR user asked me the following question:

> After fitting the age-length data into VBGM, I overviewed the results. But I can't find the coefficient of determination ($$latex r^2$$) for the VBGM fitting. Because some reviewer want the the coefficient of determination, I have to show it.

In general, the traditional ``coefficient of determination'' is not defined for non-linear regressions models.  To quote Douglas Bates from this [R-Help thread](https://stat.ethz.ch/pipermail/r-help/2002-July/023461.html) ...

> There is a good reason that an nls model fit in R does not provide
> r-squared - r-squared doesn't make sense for a general nls model.

> One way of thinking of r-squared is as a comparison of the residual sum of
> squares for the fitted model to the residual sum of squares for a trivial
> model that consists of a constant only.  You cannot guarantee that this is a
> comparison of nested models when dealing with an nls model.  If the models
> aren't nested this comparison is not terribly meaningful.

> So the answer is that you probably don't want to do this in the first place.

This [Stack Exchange thread](http://stackoverflow.com/questions/14530770/calculating-r2-for-a-nonlinear-model) also points to [this article](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2892436/) as a reference.

At this point, I would "argue" with the referee about including an $r^2$ calculation.  If the referee only wants a measure of "fit" then I would probably just include a plot of the original data with the best-fit non-linear model super-imposed.

If the referree will not give in on this point then it is possible to exploit the usual definition of $r^2$ to develop what some call "quasi-$r^2$" values.  The code below demonstrates two ways to compute the two quasi-$r^2$ values mentioned in the Stack Exchange thread when applied to a von Bertalanffy growth model.

First, fit the model as demonstrated in the [Von Bertalanffy Growth Model (Intro) Vignette](https://5c3dc6c1-a-62cb3a1a-s-sites.googlegroups.com/site/fishrfiles/gnrl/VonBertalanffy.pdf) by loading the required packages ...

```{r results='hide', warning=FALSE, message=FALSE}
library(FSA)
library(FSAdata)
```

... getting the data ...

```{r}
data(Croaker2)
crm <- Subset(Croaker2,sex=="M")
```

... and fitting the model ...

```{r}
svTypical <- vbStarts(tl~age,data=crm)
vbTypical <- tl~Linf*(1-exp(-K*(age-t0)))
fitTypical <- nls(vbTypical,data=crm,start=svTypical)
summary(fitTypical)
```

One quasi-$r^2$ is computed as the square of the correlation between the actual "y" values (lengths in this case) and the "y" values predicted from the best-fit model.  This quasi-"r^2" is computed as

```{r}
predtl <- predict(fitTypical)
cor(crm$tl,predtl)^2
```

A second quasi-$r^2$ is computed using the definition of the usual $r^2$ ...

\[ r^2 = 1-\frac{SS_{Error}}{SS_{Total}} \]

where $SS_{Error}$ is most easily computed as the sum of the squared residuals from the model fit and $SS_{Total}$, in this case, is best computed as the variance of the original "y" variable times $n-1$.  Thus, this quasi-$r^2$ is computed as

```{r}
SSE <- sum(residuals(fitTypical)^2)
SST <- var(crm$tl)*(length(crm$tl)-1)
1-SSE/SST
```

*It is not clear to me that these two calculations will always come out the same as they did here (see some of the comments in the Stack Exchange thread).*

Finally, again, I think that the plot is a more reliable summary of the model fit.

```{r VBGMfit, fig.width=4, fig.height=4}
fitPlot(fitTypical,xlab="Age",ylab="Total Length (mm)",main="")
```