```{r postblog, eval=FALSE, echo=FALSE}
## Run these commands when you are ready to push out to the fishR blog.
title <- "Test"
cats <- c('Administration')
keys <- c('Administration')
setwd("C:/aaaWork/Web/fishR/BlogExamples")
source('WPSetup.R')
knit2wp("Test.Rmd",publish=FALSE,shortcode=FALSE,
        title=title,categories=cats,mt_keywords=keys)
```

```{r setup, echo=FALSE, results='hide'}
# Having this separate from the setup file allows one to knit this in RStudio.
opts_chunk$set(dev="png",tidy=FALSE)
```

## A Question ##
Recently, a student, working on her Senior Thesis at [Northland College](http://www.northland.edu), asked me the following question:

> Attached is an Excel file with three "important to R" worksheets.  The only thing that connects all 3 worksheets is the Lift.ID variable.  Is there a way to tell R this so that I don't have to paste all three worksheets onto one new worksheet and then manually format the data so that there is one Lift.ID that connects the Count sheet info to the Lift sheet info?

## An Answer ##
Or course, the best answer to this (type of) question would be to teach this student how to use a relational database and generate a query.  Getting that into our curriculum is a discussion for another time (and venue).

However, as these data were fairly simple, her question can be completed in R with `merge()`.  I demonstrate this below with two practice data frames that each have a Lift.ID variable.  To test behavior, `df2` only had one species for `lift.ID==5` and neither species for `lift.ID==6`.  The data frames were constructed with (again, these are just example data, real data would be imported from an external file)

```{r}
( df1 <- data.frame(Lift.ID=1:6,
                    eff=c(3,4,2,5,1,1),
                    loc=c("A","B","A","C","B","D")) )
( df2 <- data.frame(Lift.ID=c(rep(1:4,each=2),5),
                    species=c(rep(c("LKT","PWF"),4),"LKT"),
                    catch=round(rnorm(9,mean=20,sd=3),0)))
```