---
layout: page
title: "Fisheries Data by Package"
---

<!-- This is meant to be run with aaaCompile.R, not as a standalone -->

Fisheries-related data sets in the **FSA** (v`r packageVersion("FSA")`), **FSAdata** (v`r packageVersion("FSAdata")`), and **fishmethods** (v`r packageVersion("fishmethods")`) packages are listed below by package.  The data sets in **FSA** and **FSAdata** may be viewed with ![view](`r iview`), downloaded (as a CSV file) with ![download](`r idown`), and the meta-information documentation may be viewed with ![documentation](`r idoc`).

```{r echo=FALSE, results='asis'}
## Cycle through the packages
for (i in pkgs) {
  ## Get all data sets in the package
  res <- help.search(keyword="datasets",package=i)$matches[,c("Package","Topic","Title")]
  ## Remove the data sets that have lit in the name (PSDlit and WSlit)
  res <- res[!grepl("lit",res$Topic),]
  ## Make a heading
  cat("----\n\n")
  cat("##",i,"\n")
  ## Make an unnumbered list of the data sets
  for (j in 1:nrow(res)) {
    tmp <- paste0("* **",res[j,"Topic"],"** -- ")
    tmp <- paste0(tmp,"[![view](",iview,")](",gh1,res[j,"Package"],gh2,res[j,"Topic"],".csv)  ")
    tmp <- paste0(tmp,"[![download](",idown,")](",rgh1,res[j,"Package"],rgh2,res[j,"Topic"],".csv)  ")
    tmp <- paste0(tmp,"[![documentation](",idoc,")](data-html/",res[j,"Topic"],".html) -- ")
    tmp <- paste0(tmp,"*",res[j,"Title"],"*\n")
    cat(tmp)
  }
  ## Add a space before the next heading
  cat("\n")  
}
```

----

## fishmethods

```{r echo=FALSE, results='asis'}
## Get all data sets in the package
res <- help.search(keyword="datasets",package="fishmethods")$matches[,c("Package","Topic","Title")]
## Make an unnumbered list of the data sets
for (j in 1:nrow(res)) {
  tmp <- paste0("* **",res[j,"Topic"],"** -- ")
  tmp <- paste0(tmp,"*",res[j,"Title"],"*\n")
  cat(tmp)
}
```
